---
title: "Northeast Regional COVID-19 cases "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Northeast Regional COVID-19}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Overview
This package pulls daily data from the Johns Hopkins University [COVID-19 repository](https://github.com/CSSEGISandData/COVID-19/), and adapts code from Rami Krispin's [coronavirus](https://github.com/RamiKrispin/coronavirus) R package, which has a very nice dashboard. 

This one is designed to focus on specific regions, with a particular interest in Worcester, MA, so we can keep the Clark University community informed about the rate of new cases in our immediate vicinity. We are also trying to provide finer spatial resolution by reading in county-level cases reported by state health authorities in the region.

## Data prep
We use the JHU case to provide statewide totals, and the county level data scraped from particular states for mapping cases at county scales. State-level data are used for time series.

```{r, message=FALSE, warning=FALSE}
library(tidyverse)
library(maps)
library(sf)
library(patchwork)
library(covid19clark)

# county and state maps
data("counties", package = "covid19clark")
data("states", package = "covid19clark")

# case data
f <- system.file("extdata/covid19_ts.csv", package = "covid19clark")
jhu_cases <- read_csv(f) %>% filter(country == "US") %>%
  mutate(county = ifelse(grepl(",", admin), gsub(",.*", "", admin), NA)) %>%
  mutate(state = ifelse(grepl(",", admin), gsub(".*, ", "", admin), admin)) %>%
  dplyr::select(admin, state, county, date, cases)

# vector of state names for matching
state_names <- tibble(state1 = state.name, state2 = state.abb) %>%
  mutate(state1 = tolower(state1))

# state cases
state_cases <- jhu_cases %>% filter(is.na(county)) %>%
  mutate(state = tolower(state)) %>% dplyr::select(-admin) %>%
  left_join(., as_tibble(states) %>% dplyr::select(-geom)) %>%
  rename(state1 = state) %>%
  left_join(., state_names) %>%
  dplyr::select(state1, state2, county, x, y, date, cases)

# county cases
f <- system.file("extdata/covid19_daily.csv", package = "covid19clark")
daily_cases <- read_csv(f) %>% filter(ctry == "US") %>% 
  rename(state = prov, county = admin) %>% 
  mutate(county = tolower(county), state = tolower(state)) %>% 
  dplyr::select(state, county, x, y, date, cases, deaths, recovered)
```

## Regionalization
We define a focal city (Worcester), and then figure out which states fall within varius radii around Worcester. These are the basis for time series analyses. 
```{r, message=FALSE, warning=FALSE}
# select city
focal_city <- focal_point("Worcester", "USA")

# buffer and focal city
city_buffers <- lapply(1:4, function(x) focal_city %>% st_buffer(dist = x))
 
# extract region: neighboring states intersecting buffers
states_in_buffers <- lapply(city_buffers, function(x) {
  ind <- unlist(st_intersects(x, states))
  as_tibble(states) %>% slice(ind) %>% pull(state)
})
```

## Plots
```{r, message=FALSE, warning=FALSE}
# function to define breaks for bubble legend
case_breaks <- function(x) {
  size_brks <- seq(min(x), max(x), round(max(x) / 5))
  if(max(size_brks) < max(x)) {
    size_brks <- c(size_brks, max(x))
  } else{
    size_brks
  }
  return(size_brks)
}

# plots
# plot 1: County case map, for counties within 200 km radius
daily_cases_region <- daily_cases %>%
  filter(state %in% states_in_buffers[[2]])

daily_cases_region %>% filter(cases == max(cases))
# plotting parameters
size_brks <- case_breaks(daily_cases_region %>% pull(cases))  # breaks
# valid_states <- unique(county_cases$state1)
# states <- states %>%
#   mutate(Complete = ifelse(state %in% valid_states, "yes", "no"))
pal <- "RdYlBu"  # color palette
city_box <- st_bbox(city_buffers[[2]])

# plot
p1 <- ggplot(states) + geom_sf() +
  scale_fill_manual(values = c("grey50", "grey")) +
  geom_sf(data = focal_city, col = "black", shape = 3, size = 2) +
  geom_point(data = daily_cases_region, aes(x = x, y = y, size = cases),
             shape = 16, stroke = FALSE, color = "red3") +
  xlab("") + ylab("") + ggtitle("Cases reported by county") +
  scale_size_continuous(name = "Cases", limits = range(size_brks),
                        breaks = size_brks) +
  coord_sf(xlim = city_box[c(1, 3)] + c(-0.5, 0.5),
           ylim = city_box[c(2, 4)] + c(0.5, -0.5)) +
  guides(color = guide_colourbar(order = 1),
         size = guide_legend(order = 2)) +
  theme_minimal()

## plot 2
state_cases_recent <- state_cases %>% group_by(state1) %>%
  filter(state1 %in% states_in_buffers[[4]]) %>%
  filter(date == max(date))
state_box <- states %>%
  filter(state %in% states_in_buffers[[2]]) %>% st_bbox()
size_brks <- case_breaks(state_cases_recent %>% pull(cases))

p2 <- ggplot(states) + geom_sf() +
  geom_point(data = state_cases_recent, aes(x = x, y = y, size = cases),
             shape = 16, stroke = FALSE, color = "red3") +
  geom_sf(data = focal_city, col = "black", shape = 3, size = 2) +
  ggtitle("Cases reported by state") +
  scale_size_continuous(name = "Cases", limits = range(size_brks),
                        breaks = size_brks) +
  coord_sf(xlim = state_box[c(1, 3)] + c(-0.5, 0.5),
           ylim = state_box[c(2, 4)]) +
  guides(color = guide_colourbar(order = 1),
         size = guide_legend(order = 2)) +
  theme_minimal()

state_abbrevs <- lapply(states_in_buffers, function(x) {
  state_names %>% filter(state1 %in% x) %>% pull(state2)
})
case_rates <- lapply(c(1, 4), function(x) {  # x <- 4
  state_cases %>% filter(state1 %in% !!states_in_buffers[[x]]) %>%
    group_by(date) %>%
    summarize(cases = sum(cases)) %>%
    mutate(states = paste(state_abbrevs[[x]], collapse = "-")) %>%
    mutate(radius = paste0("~", x, "00 km")) %>%
    dplyr::select(radius, states, date, cases)
}) %>% do.call(rbind, .)

ylimit <- max(case_rates$cases)
p3 <- case_rates %>%
  filter(date > "2020-03-01" & radius %in% paste0("~", c(1, 4), "00 km")) %>%
  ggplot() + geom_line(aes(x = date, y = cases, color = states)) +
  ylim(0, ylimit) + ggtitle("Cases within region") +
  scale_color_manual(name = "Region", values = c("red", "blue")) +
  scale_x_date(date_labels = "%b %d", date_breaks = "4 days") +
  theme_minimal()

# rate plot, all states in 2 degree radius
pal2 <- c("purple", "blue4", "blue1", "lightblue", "orange", "red1", "red4")
p4 <- state_cases %>%
  filter(state1 %in% states_in_buffers[[2]]) %>%
  filter(date > lubridate::ymd("2020-03-01")) %>%
  ggplot() + geom_line(aes(date, cases, color = state1)) +
  ylim(0, ylimit) +
  xlab("") + ylab("Cases") +
  ggtitle("Cases per state") +
  scale_color_manual(name = "", values = pal2) + 
  scale_x_date(date_labels = "%b %d", date_breaks = "4 days") +
  theme_minimal()

op <- (p1 + p2) / (p3 + p4)
f <- here::here("vignettes/figures/case_maps.png")
ggsave(op, filename = f, height = 7, width = 10, dpi = 300)
```

```{r, echo=FALSE, out.width='100%'}
knitr::include_graphics(f)
```

