---
title: "Northeast Regional COVID-19 cases "
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Northeast Regional COVID-19}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(tidyverse)
library(covid19clark)
library(sf)
```

## Overview
This package pulls the daily time series data from the Johns Hopkins University [COVID-19 repository](https://github.com/CSSEGISandData/COVID-19/), and adapts code from Rami Krispin's [coronavirus](https://github.com/RamiKrispin/coronavirus) R package, which has a very nice dashboard. 

This one is designed to focus on specific regions, with a particular interest in Worcester, MA, so we can keep the Clark University community informed about the rate of new cases in our immediate vicinity. We are also trying to provide finer spatial resolution by reading in county-level cases reported by state health authorities in the region.

## Data
```{r, message=FALSE, warning=FALSE}
cases <- get_jhu_ts2()
us_states <- sf::st_read(
  system.file("extdata/us_states.geojson", package = "covid19clark"),
  quiet = TRUE
) %>% mutate(state = as.character(state))

# US counties
us_counties <- sf::st_read(
  system.file("extdata/us_counties.geojson", package = "covid19clark"),
  quiet = TRUE
)
```


## Regionalization
We define a focal city (Worcester), and then figure out which states fall within varius radii around Worcester. These are the basis for time series analyses.
```{r, message=FALSE, warning=FALSE}
# select city
focal_city <- focal_point("Worcester", "USA")

# buffer and focal city
city_buffers <- lapply(seq(100, 400, 100) * 1000, function(x) {
  buff <- st_transform(focal_city, crs = 102008) %>% st_buffer(dist = x)
  buff <- st_transform(buff, crs = 4326)
})

# extract region: neighboring states intersecting buffers
states_in_buffers <- lapply(city_buffers, function(x) {
  ind <- unlist(st_intersects(x, us_states))
  as_tibble(us_states) %>% slice(ind) %>% pull(state)
})
```

## Data prep
Prepare datasets first
```{r, message=FALSE, warning=FALSE}
# data
# cases %>% filter(state == "New York" & date == max(date)) %>% View()
cases <- cases %>% group_by(combined_key) %>%
  mutate(newcases = cases - lag(cases), newdeaths = deaths - lag(deaths)) %>%
  mutate(newdeaths = ifelse(newdeaths < 0, 0, newdeaths)) %>%
  mutate(case_rate = cases / (population / 1000),
         newcase_rate = newcases / (population / 1000),
         death_rate = deaths / (population / 1000),
         newdeath_rate = newdeaths / (population / 1000)) %>%
  select(county, state, combined_key, x, y, date, population,
         cases, newcases, case_rate, newcase_rate,
         deaths, newdeaths, death_rate, newdeath_rate) %>%
  ungroup()

state_names <- tibble::tibble(state = state.name, stateabb = state.abb) %>%
  mutate(state = tolower(state)) %>%
  bind_rows(., tibble(state = "district of columbia", stateabb = "DC"))

state_cases <- cases %>% group_by(state, date) %>%
  summarize(cases = sum(cases), deaths = sum(deaths, na.rm = TRUE),
            population = sum(population)) %>%
  mutate(newcases = cases - lag(cases), newdeaths = deaths - lag(deaths)) %>%
  mutate(newcases = ifelse(newcases < 0, 0, newcases),
         newdeaths = ifelse(newdeaths < 0, 0, newdeaths)) %>%
  mutate(case_rate = cases / (population / 1000),
         newcase_rate = newcases / (population / 1000),
         death_rate = deaths / (population / 1000),
         newdeath_rate = newdeaths / (population / 1000)) %>%
  select(state, date, cases, newcases, case_rate, newcase_rate,
         deaths, newdeaths, death_rate, newdeath_rate) %>%
  ungroup()

state_cases <- left_join(
  state_cases %>% mutate(state = tolower(state)),
  state_names %>% mutate(stateabb = tolower(stateabb))
) %>% select(state, stateabb, date, !!names(.))
```

## Plots

### Plot parameters
```{r, message=FALSE, warning=FALSE}
case_breaks <- function(x, min_cases, base = 5, nbreaks = 5) {
  rng <- c(0.0, max(x, na.rm = TRUE) + (base - max(x, na.rm = TRUE) %% base))
  brks <- seq(rng[1], rng[2], diff(rng) / nbreaks)
  brks[1] <- min_cases
  return(brks)
}

# pal <- RColorBrewer::brewer.pal(9, "Reds")
case_legend <- function(x) {
  x +
  annotate("rect", xmin = -69.3, xmax = -67, ymin = 41.7, ymax = 42.5,
           fill = "grey") +
  annotate("text", x = -68.9, y = 42.3, label = "Deaths", hjust = 0) +
  annotate("point", x = -69.1, y = 42.3, color = "red", shape = 16, size = 3) +
  annotate("text", x = -68.9, y = 41.9, label = "Cases", hjust = 0) +
  annotate("point", x = -69.1, y = 41.9, color = "yellow", shape = 1,
           size = 3)
}

# # plotting parameters
cases_now <- cases %>% filter(date == max(date)) %>%
  filter(!grepl("out of|unassigned", county, ignore.case = TRUE))
size_brks <- cases_now %>% pull(case_rate) %>% .[na.omit(.)] %>%
  case_breaks(., min_cases = 0.01, base = 5, nbreaks = 5) %>% round(., 2)
# size_brks <- round(case_breaks(cases %>% pull(case_rate) %>% .[na.omit(.)]), 2)
city_box <- st_bbox(city_buffers[[3]])

# cases %>% filter(date > "2020-06-28") %>% filter(state == "Rhode Island") %>% View()

```

### Case map (rate)
```{r, message=FALSE, warning=FALSE}
p1_base <- ggplot(us_counties) + geom_sf(fill = "grey80", lwd = 0.3) +
  geom_sf(data = focal_city, col = "blue1", shape = 3, size = 2) +
  geom_sf(data = us_states, col = "black", fill = "transparent")

p1 <- p1_base +
  geom_point(data = cases_now, aes(x = x, y = y, size = case_rate),
             shape = 1, color = "yellow") +
  geom_point(data = cases_now, aes(x = x, y = y, size = death_rate),
             shape = 16, stroke = FALSE, color = "red") + xlab("") + ylab("") +
  ggtitle("Case and deaths per 1000 people") +
  scale_size_continuous(name = "N per 1000", limits = range(size_brks),
                        breaks = size_brks) +
  coord_sf(xlim = city_box[c(1, 3)] + c(-0.1, 0.1),
           ylim = city_box[c(2, 4)] + c(0.1, -0.1)) + theme_minimal() +
  theme(legend.key = element_rect(fill = "grey", linetype = 0))
p1 <- p1 %>% case_legend(.)
```

### Case map (total)
```{r, message=FALSE, warning=FALSE}
# cases_now %>% filter(state == "New York") %>% View()
size_brks <- cases_now %>% pull(cases) %>%
  case_breaks(., 1, base = 1000, nbreaks = 5) %>% round(., 2)
size_brks <- c(size_brks[1], 100, 1000, 10000, 50000, 100000,
               size_brks[length(size_brks)])
p2 <- p1 +
  geom_point(data = cases_now, aes(x = x, y = y, size = cases), shape = 1,
             color = "yellow") +
  geom_point(data = cases_now, aes(x = x, y = y, size = deaths),
             shape = 16, stroke = FALSE, color = "red") + xlab("") + ylab("") +
  ggtitle("Total cases and deaths") +
  scale_size_continuous(name = "N", limits = c(0.001, max(size_brks)),
                        breaks = size_brks) +
  coord_sf(xlim = city_box[c(1, 3)] + c(-0.1, 0.1),
           ylim = city_box[c(2, 4)] + c(0.1, -0.1)) +
  theme_minimal() +
  theme(legend.key = element_rect(fill = "grey", linetype = 0))
p2 <- p2 %>% case_legend(.)
```

### Cases per state (region)
```{r, message=FALSE, warning=FALSE}
# rate plot, all states in 2 degree radius
ylimit <- max(state_cases$cases)
theme_mods <- theme(
  legend.position = "none",
  axis.text.x = element_text(size = 8, angle = 90, vjust = 0.5)
)
pal2 <- c("purple", "blue4", "blue1", "lightblue", "orange", "red1", "red4")
dt_brks <- seq.Date(as.Date("2020-03-01"), max(cases$date), by = "month")
dt_brks[length(dt_brks)] <- max(cases$date)

# log10y <- function(x) paste0(round_any(10^x / 1000, 0.01) , "K")

p3_dat <- state_cases %>% filter(state %in% states_in_buffers[[2]]) %>%
  filter(date > lubridate::ymd("2020-03-01")) %>%
  mutate(stateabb = toupper(stateabb))

dt_rng <- range(p3_dat$date)
p3 <- ggplot(p3_dat) + geom_line(aes(date, cases, color = stateabb)) +
  facet_wrap(vars(stateabb), nrow = 1) +
  xlab("") + ylab("N cases") +
  ggtitle("Cases per state") +
  scale_color_manual(name = "", values = pal2) +
  scale_y_continuous(trans = "log10", labels = as.integer,
                     breaks = 10^(0:6), limits = c(1, 10^6), expand = c(0,0)) +
  scale_x_date(date_labels = "%b %d", breaks = dt_brks) +
  theme_minimal() + theme_mods

```

### Daily cases per state
```{r, message=FALSE, warning=FALSE}
ys <- c(0, 0.48, 1, 1.48, 2, 2.48, 3, 3.48, 4, 4.48, 5)
ylimit <- max(p3_dat$newcases)
p3a <- ggplot(p3_dat) +
  geom_line(aes(date, newcases, color = stateabb), size = 0.5) +
  facet_wrap(vars(stateabb), nrow = 1) + xlab("") + ylab("N cases") +
  ggtitle("Daily cases per state") + #ylim(0, ylimit) +
  scale_color_manual(name = "", values = pal2) +
  scale_y_continuous(trans = "log10", labels = as.integer, breaks = 10^(0:4),
                     limits = c(1, ylimit), expand = c(0,0)) +
  scale_x_date(date_labels = "%b %d", breaks = dt_brks) +
  theme_minimal() + theme_mods
```

### Deaths per state
```{r, message=FALSE, warning=FALSE}
ylimit2 <- max(state_cases$deaths)
p4 <- p3_dat %>%
  ggplot() + geom_line(aes(date, deaths, color = stateabb)) +
  facet_wrap(vars(stateabb), nrow = 1) + xlab("") + ylab("N deaths") +
  ggtitle("Cumulative deaths per state") +
  scale_color_manual(name = "", values = pal2) +
  scale_y_continuous(trans = "log10", labels = as.integer,
                     breaks = 10^(0:5), limits = c(1, 10^5), expand = c(0,0)) +
  scale_x_date(date_labels = "%b %d", breaks = dt_brks) +
  theme_minimal() + theme_mods
```


### Daily deaths per state
```{r, message=FALSE, warning=FALSE}
ylimit2 <- max(state_cases$newdeaths)
p4a <- p3_dat %>%
  ggplot() + geom_line(aes(date, newdeaths, color = stateabb), size = 0.5) +
  facet_wrap(vars(stateabb), nrow = 1) + xlab("") + ylab("N deaths") +
  ggtitle("Daily deaths per state") +
  scale_color_manual(name = "", values = pal2) +
  scale_y_continuous(trans = "log10", labels = as.integer,
                     breaks = 10^(0:4), limits = c(1, 10^4), expand = c(0,0)) +
  scale_x_date(date_labels = "%b %d", breaks = dt_brks) +
  theme_minimal() + theme_mods
```

```{r, echo=FALSE, message=FALSE, warning=FALSE}
# library(cowplot)
# theme_set(theme_cowplot(font_size = 10))
# theme_mod <- theme(axis.text = element_text(size = 8),
#                    legend.text = element_text(size = 8),
#                    title = element_text(size = 8))
ggsave(plot = p1,
       filename = here::here("vignettes/figures/caserate_map.png"), width = 7,
       height = 6, units = "in", dpi = 300)
ggsave(plot = p2,
       filename = here::here("vignettes/figures/case_map.png"), width = 7,
       height = 6, units = "in", dpi = 300)
ggsave(plot = p3,
       filename = here::here("vignettes/figures/statecases_cumulative.png"),
       width = 9, height = 3, units = "in", dpi = 300)
ggsave(plot = p3a,
       filename = here::here("vignettes/figures/statecases_daily.png"),
       width = 9, height = 3, units = "in", dpi = 300)
ggsave(plot = p4,
       filename = here::here("vignettes/figures/statedeaths_cumulative.png"),
       width = 9, height = 3, units = "in", dpi = 300)
ggsave(plot = p4a,
       filename = here::here("vignettes/figures/statedeaths_daily.png"),
       width = 9, height = 3, units = "in", dpi = 300)
```

```{r, echo=FALSE, out.width='95%', fig.show='hold', fig.align='center'}
figs <- c(here::here("vignettes/figures/caserate_map.png"), 
          here::here("vignettes/figures/case_map.png"))
knitr::include_graphics(figs) 
knitr::include_graphics(
  here::here("vignettes/figures/statecases_daily.png")
) 
knitr::include_graphics(
  here::here("vignettes/figures/statecases_cumulative.png")
) 
knitr::include_graphics(
  here::here("vignettes/figures/statedeaths_daily.png")
) 
knitr::include_graphics(
  here::here("vignettes/figures/statedeaths_cumulative.png")
) 

```



